Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-12 manuscript/response_to_reviews.Rmd

[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)

I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.

[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

**We have *also* added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**

@@

- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?

**It's not unprecedented to see some trees exhibit increased growth during drought, and not surprising. Stand dynamics and other factors result in a lot of scatter in the relationship between tree growth and climate, so while the distribution of $Rt$ in Fig. 1b was shifted left (normal year should be normally distributed around 1), there still are some trees with increased growth. Our analysis helps to work out which trees these were (often shorter trees).**

**We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought, and consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al. 2015). It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation"**

**We have also added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**

@@

[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?

*Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *

**To address the question about species, we have added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
work in new species variation

*Community-level drought responses*

@@

@@

Larger-diameter trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually, although DBH was not significant during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ in single-variable tests (Tables 1, 4). When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two models for 1999 (Tables 1, 5).

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **2d**). When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$. Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).  When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed trees the highest. In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. **2**). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. **2a**). Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. **2b**). Air temperature did not vary across the vertical profile (Fig. **2c**).

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015]. Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests. The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].

>>>>

*Community-level and species' drought responses*

@@

Responses varied across species and by drought (Fig. **2**). 
Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **~0.6**) and highest in *Fagus grandifolia* (mean $Rt$ = **~0.9**).

@@

Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually, although $ln[H]$ was not significant during 1977 or 1999 individually (Tables 1, 4). When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two models for 1999 (Tables 1, 5).

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$. Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).  When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed trees the highest. In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. **3**). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. **3a**). Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. **3b**). Air temperature did not vary across the vertical profile (Fig. **3c**).

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **2**) and associated traits at a single site [see also @elliott_forest_2015]. Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests. The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].
